### 2024-01-22 iterative_monomorphisation.tex
updated doc

\usepackage{soul}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}

@@

TODO:
   \begin{itemize}
      \item \st{contact Sascha Boehme about authorship}

      \item it seems that dissentangling the implemntation and low-level description is quite difficult and may not be appropriate, it might be best to instead fuse low-level description and implementation and then split the section into substitution generation and clause generation

      \item are there any implementation of this (or comparable) algorithm in that have been evaluated?
      \item read mangling paper in detail
      \item finish implementation (mostly there)
      \item test and evaluate implementation (well on the way)
      \item (can wait) merge implementation (review code with active contributor(s) beforehand)
      \item find pretty inline setup for pseudo code
      \item (in order) write high/low level algo
      \item write implementation and evaluation
      \item write background
      \item write related work
      \item write conclusion
      \item write intro
      \item reread, spellcheck ...
    \end{itemize}

@@

    \item basic idea of the algorithm\\
    A given problem is seen as a set of polymorphic clauses. Type variables of the problem are quantified at the clause level. The type arguments of the problem are collected and the uninstantiated type arguments are matched against those that contain no type variables. This yields new type arguments, both uninstantitated and ground, the process is then repeated making use of the newly derived type arguments.

@@

@@

Therefore, the previous problem would never instantiate a clause, despite the problem becoming trvially contradictory if the the last clause and any of the first two are instantiated.

% target page number: 2-4
\section{Low level algorithm}

@@

\begin{algorithm}
\begin{algorithmic}[1]
\Function{Type arguments}{Literal: lit}

@@

\begin{algorithm}
\begin{algorithmic}[1]
\Function{Initialisation}{Problem: P}
   \ForAll{clause \(C \in P\)}
      \ForAll{literal \(lit \in C\)}
         \State let fun\_sym\_ty\_args = Type arguments(lit)
         \ForAll{(fun\_sym, ty\_args) in fun\_sym\_ty\_args}
         \If{\( \forall ty\_arg \in ty\_args. is\_ground(ty\_arg)\) }

@@

\begin{algorithm}
\begin{algorithmic}[1]
\Function{Iteration}{Problem: P, U, I, S}
   \ForAll {clause \(C \in P\)}
      \ForAll {\(fun\_sym, ty\_args \in U(C)\)}
      \ForAll {\( ty\_args' \in I(fun\_sym) \)}

@@

A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached. Then substitutions can be discarded depending on whether they instantiate or not a given set of type arguments

@@

\begin{algorithm}
\begin{algorithmic}[1]
\Function{Instantiate clause}{Clause: C, Substitutions S}

@@

\begin{algorithm}
\begin{algorithmic}[1]
\Function{Generate clauses}{Problem: P, Substitutions: S}

@@

\State \Return \(P\)

@@

% target page number: 2-4
\section{Implementation and evaluation}

@@

Compare the following:
    \begin{itemize}
        \item no call to e (no monomorphisation) (using portfolio mode, this serves as reference)
        \item call e on start
        \item call e (default) monomorphisation (this is too late)
        \item call e at 0.05 * timeout
        \item call e at 0.1 * timeout (may still be too late)
        \item call e at some time before 0.05 * timeout ?
        \item different bounds worth trying: give up on the 2000 clause maximum
              add a floor for polymorphic type args and also a ceiling 
              messa around with multipliers a little bit
         \item perhaps look into some dynamic bounds
         \item consider calling a different first order prover
    \end{itemize}

Compare performance with implementation of a similar algorithm (if any).

Maybe compare with implementation of other kinds of monomorphisation algorithms (if appropriate). This would require implementing a guards or tags method which could take a while

Comments on results:
So far from the testing i've done:
\begin{itemize}
   \item we gain very few problems compared to portfolio mode, this is very bad especially since we only look at polymorphic problems, where this algorithm should shine at least in proportional terms
   \item these can easily be lost with small bound adjustments, this isn't great but not catastriphic
   \item almost all newly gained problems are solved with the 0.0 e call, this one is catastrophic
\end{itemize}

@@

\begin{itemize}
    \item that part of Sascha Boehme's thesis (Proving Theorems of Higher-Order Logic with SMT Solvers)
    \item How to Encode Polymorphic Types Safely and Efﬁciently
    \item Encoding Monomorphic and Polymorphic Types
\end{itemize}

@@

% summary 
In conclusion, this paper has given a theoretical and practical description of an iterative monomorphisation algorithm. The practical description gives some insights on how it may be efficiently implemented. Additionally, its implementation in zipperposition demonstrates its practical feasibility.

However, the evaluation of this implementation against the TPTP benchmarks show dissapointing results in light of the efficiency of similar ideas in other frameworks. Indeed, zipperposition can only solve a handful of additional problems thanks to monomorphisation. Despite these issues, a refinment of the existing implementation paired with a better choice of bounds and options may allow for an improvement in results [decent enough to warrant having spent so much time on it]

>>>>

\usepackage{soul}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}

@@

TODO:
   \begin{itemize}

      \item split the problems in two sets one for refining parameters and bounds, the other for overall evaluation
      \item read mangling paper in detail
      \item set up tables
      \item finish implementation (mostly there)
      \item test and evaluate implementation 
      \item (can wait) merge implementation (review code with active contributor(s) beforehand)
      \item find pretty inline setup for pseudo code
      \item (in order) write high/low level algo
      \item write implementation and evaluation
      \item write background
      \item write related work
      \item write conclusion
      \item write intro
      \item reread, spellcheck ...
    \end{itemize}

@@

    \item basic idea of the algorithm

    A given problem is seen as a set of polymorphic clauses. Type variables of the problem are quantified at the clause level. The type arguments of the problem are collected and the uninstantiated type arguments are matched against those that contain no type variables. This yields new type arguments, both uninstantitated and ground, the process is then repeated making use of the newly derived type arguments.

@@

[Start out by mentionning popular type encodings, such as tags and gards, explain the interest of this method]
Why we didn't implement the types / guards method: in part because it relies on inferences on types that are given in Isabelle but we would have had to do ourselves

@@

The previous problem would therefore never instantiate a clause, despite the problem becoming trvially contradictory if the the last clause and any of the first two are instantiated.

Furthermore, the finite number of iterations of the algorithm consitutes an additional source of incompleteness in practice.
[find example that would require a list list list int to be solved but only allow such a type to be generated after 3 iterations]
% target page number: 2-4
\section{Low level algorithm}

@@

\begin{algorithm}[tbh]
\begin{algorithmic}[1]
\Function{Type arguments}{Literal: lit}

@@

\begin{algorithm}[tbh]
\begin{algorithmic}[1]
\Function{Initialisation}{Problem: P}
   \ForAll{clause \(C \in P\)}
      \ForAll{literal \(lit \in C\)}
         \State let fun\_sym\_ty\_args = Type arguments(lit)
         \ForAll{(fun\_sym, ty\_args) in fun\_sym\_ty\_args}
         \If{\( \forall ty\_arg \in ty\_args. is\_ground(ty\_arg)\) }

@@

This function is very basic and simply gathers all the type arguments of the various function symbols of the problem and sorts them depending on whether they are ground or not.

\begin{algorithm}[tbh]
\begin{algorithmic}[1]
\Function{Iteration}{Problem: P, U, I, S}
   \ForAll {clause \(C \in P\)}
      \ForAll {\(fun\_sym, ty\_args \in U(C)\)}
      \ForAll {\( ty\_args' \in I(fun\_sym) \)}

@@

A consequences of these bounds is that in order to avoid unecessary computations, the program will iteratively apply substitutions until either the ground or non-ground bound is reached. Then substitutions can be discarded depending on whether they instantiate a given set of type arguments or not.

@@

\begin{algorithm}[tbh]
\begin{algorithmic}[1]
\Function{Instantiate clause}{Clause: C, Substitutions S}

@@

\begin{algorithm}[tbh]
\begin{algorithmic}[1]
\Function{Generate clauses}{Problem: P, Substitutions: S}

@@

\State \(P' = \emptyset \)

\ForAll {clause \(C \in P\)}
   \If{\(C\) is ground}
      \State \(P' = P' \cup \{mangle(C)\}\)
   \EndIf
\EndFor

\State \Return \(P'\)

@@

% target page number: 2-4
\section{Evaluation}

@@

Compare the following:
    \begin{itemize}
        \item e on original portfolio mode
        \item e with original portfolio mode and monomorphisation
        \item adjusted portforlio mode

        \item when e is called
        \item using zipp only as monomorphiser (passing the result to E, vampire, satallax, leo3)

        \item test different bounds
    \end{itemize}

@@

\begin{itemize}
    \item part of Sascha Boehme's thesis (Proving Theorems of Higher-Order Logic with SMT Solvers)
    \item How to Encode Polymorphic Types Safely and Efﬁciently, section 4

    \item Expressing polymorphic types in a many-sorted language
    \item Encoding Monomorphic and Polymorphic Types
\end{itemize}

@@

% summary
